---
title: "Function Examples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Function Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = TRUE
)
```

# Function Examples

This vignette demonstrates practical examples of using the skiResultsR functions with sample data. These examples show you exactly how to use each function and what to expect from the output.

## Setup

```{r setup}
library(skiResultsR)

# Get path to sample data with fallback for development
file_path <- system.file("extdata", "chatham_oct2023.html", package = "skiResultsR")
if (!file.exists(file_path) || file_path == "") {
  file_path <- file.path("inst", "extdata", "chatham_oct2023.html")
}

cat("Using sample file:", basename(file_path), "\n")
cat("File exists:", file.exists(file_path), "\n")
```

## Example 1: Basic Time Cleaning

This example shows how race times are cleaned and standardized:

```{r example1}
# Example time strings that might appear in race results
example_times <- c("1:23.45", "DNF", "2:15.30", "DNS", "1:45.20")

cat("Example 1: Cleaning individual race times\n")
cat("----------------------------------------\n")

# Note: clean_race_time is an internal function, but we can demonstrate the concept
for (i in seq_along(example_times)) {
  cat(sprintf("Original: %s\n", example_times[i]))
  
  # Simple time parsing demonstration
  if (toupper(example_times[i]) %in% c("DNF", "DNS")) {
    cat(sprintf("  Status: %s\n", toupper(example_times[i])))
    cat("  Cleaned: NA\n")
  } else if (grepl(":", example_times[i])) {
    parts <- strsplit(example_times[i], ":")[[1]]
    if (length(parts) == 2) {
      minutes <- as.numeric(parts[1])
      seconds <- as.numeric(parts[2])
      total_seconds <- minutes * 60 + seconds
      cat("  Status: FINISHED\n")
      cat(sprintf("  Cleaned: %.2f seconds\n", total_seconds))
    }
  }
  cat("----------------------------------------\n")
}
```

## Example 2: Processing a Complete Event

Extract and examine complete event data:

```{r example2}
if (!file.exists(file_path)) {
  cat("Sample file not available - skipping event processing example\n")
} else {
  cat("Example 2: Processing a complete event\n")
  cat("----------------------------------------\n")
  
  # Extract complete event data
  event_data <- get_event(file_path)
  event <- event_data[[1]]  # Get the inner element with class "skiResults_event"
  
  cat("Event Title:", event$event_dtls$title, "\n")
  cat("Number of races found:", length(event$races), "\n")
  if (!is.null(event$racers)) {
    cat("Number of unique racers:", nrow(event$racers), "\n")
  } else {
    cat("Number of unique racers: 0\n")
  }
  if (!is.null(event$points)) {
    total_points <- sum(sapply(event$points, function(x) if (!is.null(x)) nrow(x) else 0))
    cat("Number of points entries:", total_points, "\n")
  } else {
    cat("Number of points entries: 0\n")
  }
  if (!is.null(event$clubs)) {
    cat("Number of clubs:", nrow(event$clubs), "\n")
  } else {
    cat("Number of clubs: 0\n")
  }
}
```

## Example 3: Working with Individual Races

Extract and analyze individual race data:

```{r example3}
if (!file.exists(file_path)) {
  cat("Sample file not available - skipping race analysis example\n")
} else {
  cat("Example 3: Working with individual races\n")
  cat("----------------------------------------\n")
  
  # Get all races
  races <- get_races(file_path)
  race_ids <- names(races)
  
  cat("Available races:\n")
  for (i in seq_along(race_ids)) {
    cat(sprintf("%d. %s\n", i, race_ids[i]))
  }
  
  if (length(race_ids) > 0) {
    # Get the first race as a clean data frame
    first_race <- get_race(file_path, race_ids[1])
    
    cat(sprintf("\nAnalyzing race: %s\n", race_ids[1]))
    cat(sprintf("Number of participants: %d\n", nrow(first_race)))
    cat(sprintf("Number of columns: %d\n", ncol(first_race)))
    cat("Column names:", paste(names(first_race), collapse = ", "), "\n")
    
    # Show first few rows
    if (nrow(first_race) > 0) {
      cat("\nFirst few participants:\n")
      print(head(first_race, 3))
    }
    
    # Analyze time data if available (look for "Overall Time" or similar columns)
    time_col <- NULL
    for (col in c("Overall Time", "overall_time", "time", "Time")) {
      if (col %in% names(first_race)) {
        time_col <- col
        break
      }
    }
    
    if (!is.null(time_col)) {
      time_values <- first_race[[time_col]]
      # Try to convert to numeric
      time_numeric <- suppressWarnings(as.numeric(time_values))
      finished_times <- time_numeric[!is.na(time_numeric)]
      
      if (length(finished_times) > 0) {
        cat("\nRace Statistics:\n")
        cat("- Participants with times:", length(finished_times), "\n")
        cat("- Fastest time:", min(finished_times), "\n")
        cat("- Slowest time:", max(finished_times), "\n")
        cat("- Average time:", round(mean(finished_times), 2), "\n")
      }
      
      # Count DNS, DNF, DSQ
      time_char <- as.character(time_values)
      dns_count <- sum(grepl("^DNS$", time_char, ignore.case = TRUE))
      dnf_count <- sum(grepl("^DNF$", time_char, ignore.case = TRUE))
      dsq_count <- sum(grepl("^DSQ$", time_char, ignore.case = TRUE))
      
      if (dns_count > 0 || dnf_count > 0 || dsq_count > 0) {
        cat("\nStatus Distribution:\n")
        cat("- DNS:", dns_count, "\n")
        cat("- DNF:", dnf_count, "\n")
        cat("- DSQ:", dsq_count, "\n")
      }
    }
  }
}
```

## Example 4: Extracting Participant Information

Get detailed information about race participants:

```{r example4}
if (!file.exists(file_path)) {
  cat("Sample file not available - skipping participant example\n")
} else {
  cat("Example 4: Extracting participant information\n")
  cat("----------------------------------------\n")
  
  # First get race IDs to use for get_racers()
  races <- get_races(file_path)
  if (length(races) > 0) {
    # Get participant information for the first race
    race_id <- names(races)[1]
    participants <- get_racers(file_path, race_id = race_id)
    
    cat(sprintf("Total participants in race %s: %d\n", race_id, nrow(participants)))
    
    if (nrow(participants) > 0) {
      # Check for profile links
      if ("Profile URL" %in% names(participants)) {
        with_links <- sum(!is.na(participants$`Profile URL`) & participants$`Profile URL` != "")
        cat("Participants with profile links:", with_links, "\n")
      }
      
      # Check for club information
      if ("Club" %in% names(participants)) {
        with_clubs <- sum(!is.na(participants$Club) & participants$Club != "")
        cat("Participants with club information:", with_clubs, "\n")
      }
      
      cat("\nFirst few participants:\n")
      print(head(participants, 5))
      
      # Show unique clubs if any
      if ("Club" %in% names(participants)) {
        unique_clubs <- unique(participants$Club[!is.na(participants$Club) & participants$Club != ""])
        if (length(unique_clubs) > 0) {
          cat("\nClubs represented:\n")
          for (club in unique_clubs) {
            cat("-", club, "\n")
          }
        }
      }
    }
  } else {
    cat("No races found in file\n")
  }
}
```

## Example 5: Getting Event Details

Extract just the event metadata:

```{r example5}
if (!file.exists(file_path)) {
  cat("Sample file not available - skipping event details example\n")
} else {
  cat("Example 5: Extracting event details\n")
  cat("----------------------------------------\n")
  
  # Get event details using get_event_dtls()
  event_details <- get_event_dtls(file_path)
  
  if (!is.null(event_details) && nrow(event_details) > 0) {
    cat("Event details found:\n")
    
    # Show main details
    if (!is.na(event_details$title)) {
      cat("Title:", event_details$title, "\n")
    }
    if (!is.na(event_details$date)) {
      cat("Date:", event_details$date, "\n")
    }
    if (!is.na(event_details$slope)) {
      cat("Venue:", event_details$slope, "\n")
    }
    if (!is.na(event_details$format)) {
      cat("Format:", event_details$format, "\n")
    }
    if (!is.na(event_details$status)) {
      cat("Status:", event_details$status, "\n")
    }
  } else {
    cat("No event details found in the file\n")
  }
}
```

## Example 6: Points Analysis

Analyze points data for a specific race:

```{r example6}
if (!file.exists(file_path)) {
  cat("Sample file not available - skipping points analysis example\n")
} else {
  cat("Example 6: Points analysis\n")
  cat("----------------------------------------\n")
  
  # First get race IDs
  races <- get_races(file_path)
  if (length(races) > 0) {
    # Get points for the first race
    race_id <- names(races)[1]
    points_data <- tryCatch({
      get_points(file_path, race_id = race_id)
    }, error = function(e) {
      # Points may not exist for all races
      NULL
    })
    
    if (!is.null(points_data) && nrow(points_data) > 0) {
      cat("Points entries found for", race_id, ":", nrow(points_data), "\n")
      
      # Show sample points data
      cat("\nSample points data:\n")
      print(head(points_data, 5))
      
      # Show column names to understand structure
      cat("\nPoints data columns:", paste(names(points_data), collapse = ", "), "\n")
    } else {
      cat("No points data found for", race_id, "\n")
      cat("(Points may not be available for all races)\n")
    }
  } else {
    cat("No races found in file\n")
  }
}
```

## Summary

These examples demonstrate the key capabilities of the skiResultsR package:

1. **Complete event extraction** with `get_event()`
2. **Individual race analysis** with `get_race()` and `get_races()`
3. **Participant management** with `get_racers()`
4. **Event metadata** with `get_event_dtls()` and `get_event_summary()`
5. **Points analysis** with `get_points()`

Each function is designed to handle real-world ski race data and provide clean, structured output for further analysis.

## Next Steps

- Try these functions with your own HTML files from skiresults.co.uk
- Combine the output with data visualization packages like `ggplot2`
- Use the structured data for statistical analysis of race performance
- Export results to CSV or other formats for sharing