---
title: "Getting Started with skiresultsR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with skiresultsR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = FALSE,
  warning = FALSE,
  message = FALSE
)
```

# Introduction

The `skiresultsR` package provides a comprehensive set of functions for extracting and analyzing ski race results from HTML files generated by skiresults.co.uk. This vignette will guide you through the main features and show you how to extract structured data from ski race events.

## Installation

```{r eval=FALSE}
# Install from GitHub
devtools::install_github("justinjtownsend/skiresultsR")

# Or install locally
devtools::install()
```

## Loading the Package

```{r setup}
library(skiresultsR)

# Get path to sample data
file_path <- system.file("extdata", "chatham_oct2023.html", package = "skiresultsR")
```

# Main Functions Overview

The `skiresultsR` package provides six main functions, organized into primary and helper functions:

## Primary Functions

- **`get_event()`** - Extract complete event data (recommended starting point)
- **`get_races()`** - Extract all race tables from an event
- **`get_race()`** - Extract a specific race by race type/ID

## Helper Functions

- **`get_racers()`** - Extract unique racer information with profile links
- **`get_points()`** - Extract racer points across all categories  
- **`get_event_summary()`** - Extract event metadata (title, date, venue, etc.)

# Basic Usage

## 1. Getting Complete Event Data

The easiest way to start is with `get_event()`, which extracts all available information:

```{r get-event}
# Extract complete event data
event_data <- get_event(file_path)

# View the structure
str(event_data, max.level = 1)
```

The `get_event()` function returns a nested list with event_id as the name. The inner element (accessed with `event_data[[1]]`) contains:

- `event_dtls`: Event details (title, date, slope, format, status)
- `race_types`: Race types information
- `races`: All race tables with complete race information
- `racers`: Unique racers with their profile links and club information
- `points`: Points data across all categories and races
- `clubs`: Club information found in the event

## 2. Exploring Event Details

```{r explore-event}
# Get the actual event object (inner element)
event <- event_data[[1]]

# Display key information from event_dtls
cat("Event Title:", event$event_dtls$title, "\n")
cat("Date:", event$event_dtls$date, "\n")
cat("Venue:", event$event_dtls$slope, "\n")

# Show available race types
cat("\nAvailable Race Types:\n")
print(event$race_types$race_type)
```

## 3. Working with Race Data

```{r work-races}
# View available races
race_names <- names(event$races)
cat("Found", length(race_names), "races:\n")
cat(paste(race_names, collapse = ", "), "\n")

# Examine the first race
if (length(race_names) > 0) {
  first_race <- event$races[[1]]
  cat("\nFirst race structure:\n")
  cat("- Race ID:", race_names[1], "\n")
  cat("- Number of participants:", nrow(first_race), "\n")
  cat("- Columns:", paste(names(first_race), collapse = ", "), "\n")
  
  # Show first few rows
  if (nrow(first_race) > 0) {
    cat("\nFirst few participants:\n")
    print(head(first_race, 3))
  }
}
```

## 4. Analyzing Racer Information

```{r analyze-racers}
# View racer information
racers <- event$racers
cat("Found", nrow(racers), "unique racers\n")

cat("\nFirst few racers:\n")
print(head(racers, 5))

# Check for club information
if ("Club" %in% names(racers)) {
  racers_with_clubs <- sum(!is.na(racers$Club) & racers$Club != "")
  cat("\nRacers with club information:", racers_with_clubs, "\n")
}
```

# Specialized Functions

## Getting Event Summary Only

If you need event summary statistics, use `get_event_summary()` with an event object:

```{r event-summary}
# Get event summary
summary <- get_event_summary(event)

# Access race summary and participation statistics
event_id <- names(summary)[1]
cat("Event ID:", event_id, "\n")
cat("\nRace Summary:\n")
print(summary[[1]]$race_summary)
cat("\nRace Participation:\n")
print(summary[[1]]$race_participation)
```

## Getting All Races

To extract all race tables:

```{r all-races}
# Get all races
all_races <- get_races(file_path)

cat("Found", length(all_races), "races\n")

# Each element is a data frame with race results
if (length(all_races) > 0) {
  first_race_id <- names(all_races)[1]
  first_race_data <- all_races[[1]]
  cat("First race ID:", first_race_id, "\n")
  cat("First race participants:", nrow(first_race_data), "\n")
  cat("First race columns:", paste(names(first_race_data), collapse = ", "), "\n")
}
```

## Getting a Specific Race

To extract just one race by its ID:

```{r specific-race}
# Get available race IDs first
races <- get_races(file_path)
race_ids <- names(races)

if (length(race_ids) > 0) {
  # Get the first race as a clean data frame
  specific_race <- get_race(file_path, race_ids[1])
  
  cat("Extracted race:", race_ids[1], "\n")
  cat("Participants:", nrow(specific_race), "\n")
  cat("Columns:", ncol(specific_race), "\n")
  
  # Show structure
  if (nrow(specific_race) > 0) {
    print(head(specific_race, 3))
  }
}
```

## Getting Race Points

To extract points information for a specific race:

```{r race-points}
# First get race IDs
races <- get_races(file_path)
if (length(races) > 0) {
  # Get points for the first race
  race_id <- names(races)[1]
  points_data <- tryCatch({
    get_points(file_path, race_id = race_id)
  }, error = function(e) {
    # Points may not exist for all races
    NULL
  })
  
  if (!is.null(points_data) && nrow(points_data) > 0) {
    cat("Found", nrow(points_data), "points entries for", race_id, "\n")
    cat("\nSample points data:\n")
    print(head(points_data, 3))
  } else {
    cat("No points data available for", race_id, "\n")
  }
}
```

## Getting Racers Only

To extract just racer information with links:

```{r racers-only}
# First get race IDs
races <- get_races(file_path)
if (length(races) > 0) {
  # Get racer information for the first race
  race_id <- names(races)[1]
  racers_only <- get_racers(file_path, race_id = race_id)
  
  cat("Found", nrow(racers_only), "racers in race", race_id, "\n")
  
  if (nrow(racers_only) > 0) {
    # Check for profile links
    if ("Profile URL" %in% names(racers_only)) {
      with_links <- sum(!is.na(racers_only$`Profile URL`) & racers_only$`Profile URL` != "")
      cat("Racers with profile links:", with_links, "\n")
    }
    
    # Show sample
    print(head(racers_only, 3))
  }
}
```

# Data Analysis Examples

## Race Performance Analysis

```{r performance-analysis}
# Get a specific race for analysis
races <- get_races(file_path)
if (length(races) > 0) {
  race_data <- get_race(file_path, names(races)[1])
  
  # Analyze time data if available (look for "Overall Time" or similar columns)
  time_col <- NULL
  for (col in c("Overall Time", "overall_time", "time", "Time")) {
    if (col %in% names(race_data)) {
      time_col <- col
      break
    }
  }
  
  if (!is.null(time_col)) {
    time_values <- race_data[[time_col]]
    # Try to convert to numeric
    time_numeric <- suppressWarnings(as.numeric(time_values))
    finished_times <- time_numeric[!is.na(time_numeric)]
    
    if (length(finished_times) > 0) {
      cat("Race Statistics:\n")
      cat("- Participants with times:", length(finished_times), "\n")
      cat("- Fastest time:", min(finished_times), "\n")
      cat("- Slowest time:", max(finished_times), "\n")
      cat("- Average time:", round(mean(finished_times), 2), "\n")
    }
    
    # Count DNS, DNF, DSQ
    time_char <- as.character(time_values)
    dns_count <- sum(grepl("^DNS$", time_char, ignore.case = TRUE))
    dnf_count <- sum(grepl("^DNF$", time_char, ignore.case = TRUE))
    dsq_count <- sum(grepl("^DSQ$", time_char, ignore.case = TRUE))
    
    if (dns_count > 0 || dnf_count > 0 || dsq_count > 0) {
      cat("\nStatus Distribution:\n")
      cat("- DNS:", dns_count, "\n")
      cat("- DNF:", dnf_count, "\n")
      cat("- DSQ:", dsq_count, "\n")
    }
  } else {
    cat("No time column found in race data\n")
  }
}
```

## Points Analysis

```{r points-analysis}
# Analyze points distribution for a specific race
races <- get_races(file_path)
if (length(races) > 0) {
  race_id <- names(races)[1]
  points_data <- tryCatch({
    get_points(file_path, race_id = race_id)
  }, error = function(e) NULL)
  
  if (!is.null(points_data) && nrow(points_data) > 0) {
    cat("Points data for", race_id, ":\n")
    cat("Total entries:", nrow(points_data), "\n")
    # Show summary
    print(head(points_data, 5))
  } else {
    cat("No points data available for analysis\n")
  }
}
```

# Best Practices

## 1. Error Handling

Always check if files exist and handle potential errors:

```{r eval=FALSE}
tryCatch({
  event_data <- get_event(file_path)
  cat("Successfully extracted event data\n")
}, error = function(e) {
  cat("Error extracting data:", e$message, "\n")
})
```

## 2. Choosing the Right Function

- Use `get_event()` for comprehensive analysis
- Use `get_event_summary()` for quick metadata checks
- Use `get_race()` when you know the specific race you want
- Use helper functions for targeted data extraction

## 3. Data Validation

Always validate your extracted data:

```{r eval=FALSE}
# Check data quality
event_data <- get_event(file_path)
event <- event_data[[1]]  # Get the inner element

# Validate racers data
if (is.null(event$racers) || nrow(event$racers) == 0) {
  warning("No racers found - check HTML structure")
}

# Validate race data
if (is.null(event$races) || length(event$races) == 0) {
  warning("No races found - check HTML structure")
}
```

# Conclusion

The `skiresultsR` package provides a comprehensive toolkit for extracting and analyzing ski race results. The functions are designed to be intuitive and consistent, making it easy to work with ski race data in R.

For more detailed information about each function, see the function documentation using `?function_name` (e.g., `?get_event`).

## Next Steps

- Explore the function reference documentation
- Try the functions with your own HTML files
- Combine the extracted data with visualization packages like `ggplot2`
- Use the data for statistical analysis of race performance
